Naturalness and Long-lived Beyond the Standard Model Particles

### 2017-07-27 tex/draft.tex
add (or fix) Andrew's comments

    All SM particles transform according to the gauged symmetry group SU(3)cxSU(2)LxU(1)Y, where SU(3)c is associated with color charge, SU(2)L is associated with weak isospin, and U(1)Y is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are 8 gluons associated with SU(3)c, three W bosons associated with SU(2)L, and 1 B boson associated with U(1)Y. The fermions are catagorized by which representation of each group they fit into. For example, SU(3)c distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and SU(2)L distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, different values of weak hypercharge further distignuish SM particles according to a linear combination of their weak isospin and electric charge \fxnote{is there some more fundamental way to describe Y?}.

    The Higgs potential has a nonzero vacuum expectation value (VEV), which spontaneously breaks SU(2)LxU(1)Y to U(1)EM while leaving SU(3)c unscathed. Once the dust settles, the B and 3 W bosons have been mixed to produce the W+, W-, Z bosons that mediate the weak force and the photon that mediates electromagnetism. 3 of the 4 generators are broken in the process, leaving 3 extra degrees of freedom that ultimately become the masses of the W+, W-, and Z bosons while the remaining symmetry is associated with electric charge. Furthermore, expanding the Higgs field about its VEV produces mass terms for the quarks and charged leptons, thus giving mass to all massive SM particles. \fxnote{probably show a little math}

    \fxnote{kinda funny wording throughout. Maybe it would be better to trade some words for math?}
    All SM particles interact with virtual particles through loop interactions. These interactions lead to corrections to SM parameters such as particle masses. It is common to write a given parameter as a linear combination of its 'bare' (uncorrected) value and its corrections. As the only fundamental scalar in the SM, the Higgs is uniquely affected by these corrections. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the Higgs bare mass and the terms that correct it.

@@

    Order($10^{32}$) corrections imply that the bare Higgs mass must be negative and also order($10^{32}$) to reproduce the experimetnally known value. In fact, the bare mass and corrections must match to 1 part in $\approx 10^{30}$. While it's entirely possible that these two numbers just so happen to match so perfectly, many \fxnote{weak} feel that such miraculous fine tuning must be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is order(TeV).

    Before digging into possible solutions to this problem\fxnote{specify}, one other question must be addressed: why doesn't this same argument apply to all SM particles? The answer comes from naturalness statement (3) above. As a fundamental scalar, the Higgs boson lacks the chiral or gauge symmetries enjoyed by the fermions and vector bosons, respectively. These two symmetries cause the symmetry of the SM to be enhanced when the fermion and vector boson masses are set to 0. \fxnote{work out fermion case?} Setting the higgs mass to 0 in no way increases the symmetry of the theory, so we have no reason to expect it to be small.

    There are three main approaches to resolving this problem\fxnote{specify}: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass. Arguments against naturalness often involve the anthropic principle, which claims that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than 3 spatial dimensions, which lowers the Planck mass itself. Finally, the dominant BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below. \fxnote{didn't mention composite higgs}

@@

\section{Supersymmetry}
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions. In its simplest form, SUSY results in one new boson for every SM fermion and one new fermion for every SM boson \fxnote{and an extra Higgs field - can I gloss over that?}. The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-1/2 superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Together the superpartners to the SM particles are known as sparticles.

@@

    In the SM and in BSM theories, LLPs result from one of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, on the other hand, are long-lived beause the proton and electron masses leave them little phase space in which to decay \fxnote{also mention that they're both weak decays?}. \fxnote{Mention neutrino, pion, kaon, some other bound state I might be forgetting?}

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models meet different LLP criteria in different ways. The most common \fxnote{based on my assumptions} SUSY LLP occurs in models with conserved R-partity, which is a proposed symmetry that protects many models from proton decay. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as significant missing energy, making MET the conventional SUSY signature in many searches.

@@

    LLPs from highly-virtual intermediate particles can occur in models such as split SUSY, where the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark. \fxnote{cite kilian} \fxnote{mention that split throws out naturalness motivation?}

    The degenerate mass spectra of other SUSY models can also give rise to LLPs. In some AMSB models \fxnote{explain AMSB?}, for example, the zino and wino are the LSP and NLSP, respectively, and are nearly degenerate in mass. The wino then becomes long-lived when its decay is suppressed by the lack of available phase space. \fxnote{cite R\&S out of this world}
    \fxnote{merge last 2 paragraphs under 'mass spectra lead to LLP' concept?}
    \fxnote{never said anything about hidden valley or dark sector stuff. Never said the word R-hadron. Never said stealth or displaced SUSY. Should I?}

\section{The CMS Detector}
\subsection{Overview}
    \fxnote{Add LHC description}  
    \fxnote{cite cms experiment at the cern lhc}
    The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from 14 TeV proton-proton collisions at the Large Hadron Collider (LHC). To meet this goal, CMS combines information from several subdetectors nested radially in and around a 4-T superconducting solenoid. All together, CMS is 21.6 m long, 14.6 m across, and weighs approximately 12500 tons. The subdetectors that make up CMS are described below.

\subsection{Tracker}
    \fxnote{cite CMS TDR, jinst 12 c02033}
    \fxnote{maybe I should trade some number listing for talking about how silicon detectors work}
    \fxnote{what about readout, cooling, etc?}
    In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions: at r $<$ 20 cm, the high particle flux \fxnote{specify} requires small pixels to keep occupancy near 0.1\%. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

    The pixel tracker, which was upgraded in winter 2016/17, makes up the innermost layer and features approximately 124 million silicon pixels, each of which covers an active area of 100 by 150 $um^2$. The pixels are spread between four barrel layers (BPIX) at 2.9 $<$ r $<$ 16 cm and 3 endcap rings (FPIX) at 29 $<$ z $<$ 52 cm. The upgraded pixel tracker should provide 4-hit tracking out to eta $<$ 2.4 and improved b-tagging over the original pixel tracker despite the increased pileup expected in 2017.

    Just outside the pixel tracker, the silicon strip tracker (SST) occupies the 20 $<$ r $<$ 116 cm and $|z|$ $<$ 282 cm region of CMS. Composed of 9.3 million silicon strips that vary from \_ to \_ in size \fxnote{look up} depending on particle flux in the region, the SST covers an active area of roughly 196 $m^2$. At each trigger, the strip output voltage is transferred via optical link to be read out by off-detector front-end electronics \fxnote{right jargon?}.

\subsection{Electromagnetic Calorimeter}
    \fxnote{cite CMS TDR}
    \fxnote{talk about readout?}
    \fxnote{More in the way of concept/how stuff works?}
    After traversing the inner tracker, a particle would next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses 61200 lead tunstate crystals in the barrel and 7324 in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (80\% of light within 25 ns), compact (radiation length = 0.89 cm), fine-grained (Moliere length = 2.2 cm), and radiation hard (up to 10 Mrad\fxnote{learn context for this number}) calorimeter. The main drawback is the relatively low light yield (30 photon / MeV), which necesitates photodetectors with intrinsic gain that work in magnetic fields \fxnote{say B-field part better}.

    The barrel section (EB) extends radially from 129 to 177 \fxnote{look up} cm and covers a pseudorapidity range of 0 to 1.479. The crystals are tapered to approximately project back to the IP but not so perfectly that likely particle trajectories align with cracks. Each crystal is approximately one Moliere radius wide and 25.8 radiation lengths deep.

@@

\subsection{Muon System}
    \fxnote{cite CMS TDR}
    \fxnote{Specifically mention CMS design specs to meet physics goals?}
    The muon system (MS), which is the outermost subdetector, detects muons with 3 varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background, muon rates, and magnetic field strength. The RPCs are useful in the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT resolution at high momenta where the tracker pT resolution declines \fxnote{quantify or show std plot?}.

@@

\subsection{Trigger}
    \fxnote{cite CMS TDR, jinst 12 c03021}
    \fxnote{Is entire detector read out at L1?}
    \fxnote{Mention fpgas in L1 or more details about HLT?}
    \fxnote{Something about particle flow?}
    \fxnote{Sloppy, but I'm getting tired}
    The trigger reduces the data writing rate from the 40 MHz collision rate to less than 1 kHz so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). L1 uses input from ECAL, HCAL, and MS that represents physics objects (e/gamma, tau, jets, MET, and mu) and uses custom electronics to reduce the rate to $\approx$100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to further reduce the rate to the desired $<$ 1 kHz.

\section{LHC Searches for Long-lived Particles}
\subsection{Overview}
    LLPs leave distinct signatures that differentiate LLP signals from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were generally designed to detect and reconstruct stable particles from prompt decays, so LLP analyses often need to stray from the traditional analysis path in terms of object reconstruction \fxnote{really reconstruction? Maybe object definition?}, background estimation, and which data tier to use \fxnote{can I say this more generally?}. This paper will focus on four complementary LLP signatures: displaced vertices, disappearing tracks, slow-moving particles, and particles that stop completely before decaying.

@@

    Slow and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. Searches for low-velocity particles use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles \fxnote{redundant}, which normally \fxnote{quantify} have beta $>$ 0.9. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

    This search uses a simplified model in the split SUSY paradigm as a benchmark. In the simplified model, gluinos are kinematically accessible while squarks are not. The relevant process, then, is pair produced gluinos that decay to a quark and virtual squark, which then decays to a quark and the LSP, which is presumed to be a neutralino \fxnote{diagram?}. Because the squark mass is so high, the gluino lives long enough to form an R-hadron \fxnote{first use. Define here if not in LLP section} with SM quarks and propogate some measureable distance away from the IP before decaying.

@@

    The authors consider two signal processes: (1) electroweak gaugino production resulting in a disappearing track, an ISR jet, and MET and (2) gluino pair production resulting in a disappearing track, four jets, and MET. Process (1) could be the only gaugino production mode at the LHC if the gluino mass is kinematically unaccessable, but process (2), which leads to chargino production in cascade decays, will be dominant if the gluino mass is within reach. The ISR jet in process (1) is not physically required; it simply provides something for the gluinos to recoil off of so the MET signal isn't lost \fxnote{My interpretation. Run it by someone who knows stuff}.

@@

    This analysis uses the muon system to look for pairs of muons that arrive out of time with bunch crossings. The main backgrounds to such a signal are cosmics, beam halo, and muon system noise. Beam halo and muon system noise are easily removed by the selection criteria, but distinguishing signal from cosmics requires a more nuanced approach. After generating signal MC and modeling cosmics from data obtained in dedicated cosmic runs, the authors found that cosmics and signal could be differentiated using time-of-flight information from the muon system RPCs and DTs. Essentially, the direction of muon momentum could be gleaned from the timing information, and the authors required outgoing muons in both the upper and lower hemispheres.

@@

    After estimating the background by extrapolating data from the cosmic-enriched region to the signal region and determining systematic uncertainties related to the MC simulation, trigger acceptance, and luminosity, the authors calculated the expected background for LLP lifetimes ranging from 100 ns to 1 us. In all cases, the expected background was less then one event. After unblinding, the observed number of events is exactly zero in every lifetime scenerio. These results are then interpreted as upper limits on the signal production cross section \fxnote{include limit plots or table}, which are the first limits for stopped particles that decay to muons at the LHC.

@@

    No significant excess over predicted background is observed, so the results are interpreted as upper limits in the production cross section - mass plane for the considered models. By comparing the observed and theoretical limits, the authors also extract mass limits for the gluinos, stops, and staus in various models. \fxnote{Include plots - table is too big and too many numbers to list}

\section{Experimental Viability of Natural SUSY}
    When building a natural SUSY model, certain constraints are nearly unavoidable\fxnote{Dangerous. What if Linda knows how to avoid them?}. Specifically, the higgsino, which affects mH at tree level; the stop, which affects mH at one loop; and the gluino, which affets mStop at one loop and therefore mH at two, must all be heavy enough to pull mH up to 125 GeV\fxnote{Do I really understand this?} but light enough to avoid excessive fine tuning. Furthermore, the stop and gluino masses are correlated -\fxnote{m-dash?} a large difference between the two would be another source of fine tuning.

@@

\fxnote{How do LLPs open doors to naturalness and which of those doors have we already closed?}

>>>>

    All SM particles transform according to the gauged symmetry group SU(3)cxSU(2)LxU(1)Y, where SU(3)c is associated with color charge, SU(2)L is associated with weak isospin, and U(1)Y is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are 8 gluons associated with SU(3)c, three W bosons associated with SU(2)L, and 1 B boson associated with U(1)Y. The fermions are catagorized by which representation of each group they fit into. For example, SU(3)c distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and SU(2)L distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, different values of weak hypercharge further distignuish SM particles according to the Gell-Mann\fxnote{em-dash}Nishijima formula, $iY=2(Q-I3$.

    The Higgs potential has a nonzero vacuum expectation value (VEV), which spontaneously breaks SU(2)LxU(1)Y to U(1)EM while leaving SU(3)c unscathed. Once the dust settles, the B and 3 W bosons have been mixed to produce the W+, W-, Z bosons that mediate the weak force and the photon that mediates electromagnetism. \fxnote{Andrew says the next sentence isn't quite right. 3 of 4 DOF from complex Higgs field are eaten by W+,W-,Z. Remaining DOF becomes physical Higgs boson. Masses don't come from broken generators. Read about it then fix} 3 of the 4 generators are broken in the process, leaving 3 extra degrees of freedom that ultimately become the masses of the W+, W-, and Z bosons while the remaining symmetry is associated with electric charge. Furthermore, expanding the Higgs field about its VEV produces mass terms for the quarks and charged leptons, thus giving mass to all massive SM particles.\fxnote{Fermion masses ad-hoc, not fundamental, says Andrew}  \fxnote{probably show a little math}\fxnote{Why can't we add explicit mass terms to SM lagrangian w/o Higgs? Summarize section 1 of Andrew's paper}

    \fxnote{kinda funny wording throughout. Maybe it would be better to trade some words for math?}
    All SM particles interact with virtual particles through loop interactions. These interactions lead to corrections to SM parameters such as particle masses. It is common to write a given parameter as a linear combination of its 'bare' (uncorrected) value and its corrections.\fxnote{Not wrong, but sounds funny. Of course we write it that way when talking about a sum of diagrams} As the only fundamental scalar in the SM, the Higgs is uniquely affected by these corrections.\fxnote{Explain why scalars are more sensitive. Show math probably} The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the Higgs bare mass and the terms that correct it.

@@

    \fxnote{Point is that corrections come in with different signs and individual terms are huge. Really weird that they would cancel out so perfectly. Maybe add feynman diagrams to illustrate. Use Jaxodraw probably} Order($10^{32}$) corrections imply that the bare Higgs mass must be negative and also order($10^{32}$) to reproduce the experimetnally known value. In fact, the bare mass and corrections must match to 1 part in $\approx 10^{30}$. While it's entirely possible that these two numbers just so happen to match so perfectly, many \fxnote{weak} feel that such miraculous fine tuning must be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is order(TeV).

   \fxnote{Confusing, says Andrew. Maybe add math}  Before digging into possible solutions to this problem\fxnote{specify}, one other question must be addressed: why doesn't this same argument apply to all SM particles? The answer comes from naturalness statement (3) above. As a fundamental scalar, the Higgs boson lacks the chiral or gauge symmetries enjoyed by the fermions and vector bosons, respectively. These two symmetries cause the symmetry of the SM to be enhanced when the fermion and vector boson masses are set to 0. \fxnote{work out fermion case?} Setting the higgs mass to 0 in no way increases the symmetry of the theory, so we have no reason to expect it to be small.

    \fxnote{Clarify that not technically a problem. Nature might just be that way. Anthropic principle almost like an attempt to restore naturalness to a universe that appears unnatural, says Andrew} There are three main approaches to resolving this problem\fxnote{specify}: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass. Arguments against naturalness often involve the anthropic principle, which claims that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than 3 spatial dimensions, which lowers the Planck mass itself. Finally, the dominant BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below. \fxnote{didn't mention composite higgs}

@@

\section{Supersymmetry}
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions. In its simplest form, SUSY results in one new boson for every SM fermion and one new fermion for every SM boson \fxnote{and an extra Higgs field - probably should mention it. Make sure I understand it either way}. The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-1/2 superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Together the superpartners to the SM particles are known as sparticles.

@@

    In the SM and in BSM theories, LLPs result from one of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, on the other hand, are long-lived beause the proton and electron masses leave them little phase space in which to decay \fxnote{Mention that they have small couplings b/c weak force and high-mass intermediate particle is involved too}. \fxnote{Mention neutrino, pion, kaon, some other bound state I might be forgetting?}

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models meet different LLP criteria in different ways. The most common\fxnote{bad langauge. Andrew suggests 'most straightforward' or 'most ad-hoc'} SUSY LLP occurs in models with conserved R-partity, which is a proposed symmetry that protects many models from proton decay\fxnote{Be prepared to show this. Maybe add some diagrams illustrating proton decay w/o R-parity}. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as significant missing energy, making MET the conventional SUSY signature in many searches.

@@

    LLPs from highly-virtual intermediate particles can occur in models such as split SUSY, where the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark. \fxnote{cite kilian} \fxnote{mention that split throws out naturalness motivation?}\fxnote{Add diagrams}

    \fxnote{There is no zino. Maybe I meant bino? Neutralinos and charginos are the physical states that are linear combos of winos, bino, higgsinos} The degenerate mass spectra of other SUSY models can also give rise to LLPs. In some AMSB models \fxnote{explain AMSB?}, for example, the zino and wino are the LSP and NLSP, respectively, and are nearly degenerate in mass. The wino then becomes long-lived when its decay is suppressed by the lack of available phase space. \fxnote{cite R\&S out of this world}
    \fxnote{merge last 2 paragraphs under 'mass spectra lead to LLP' concept?}
    \fxnote{never said anything about hidden valley or dark sector stuff. Never said the word R-hadron. Never said stealth or displaced SUSY. Should I?} 
    \fxnote{Andrew says don't worry about mentioning models that don't come up in the searches, but be prepared to explain the basic ideas behind HV, dark sector and how they result in LLP}

\section{The CMS Detector}\fxnote{Add pictures of everything}
\subsection{Overview}
    \fxnote{Add LHC description}  
    \fxnote{Mention that ATLAS is similar}  
    \fxnote{Maybe mention PDFs. Definitely know about them}  
    \fxnote{cite cms experiment at the cern lhc}
    The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from 14 TeV\fxnote{Mention lower-energy runs} proton-proton collisions at the Large Hadron Collider (LHC). To meet this goal, CMS combines information from several subdetectors nested radially in and around a 4-T superconducting solenoid. All together, CMS is 21.6 m long, 14.6 m across, and weighs approximately 12500 tons. The subdetectors that make up CMS are described below.

\subsection{Tracker}
    \fxnote{cite CMS TDR, jinst 12 c02033}
    \fxnote{what about readout, cooling, etc?}
    \fxnote{mention that analyses below used pre-upgrade tracker}
    \fxnote{mention stereo modules and be able to explain stereo modules}
    In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices.\fxnote{out-of-fielder may not know about 'track' or 'vertex'} The tracker is subdivided into two regions: at r $<$ 20 cm, the high particle flux \fxnote{specify} requires small pixels to keep occupancy near 0.1\%. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

    The pixel tracker, which was upgraded in winter 2016/17, makes up the innermost layer and features approximately 124 million silicon pixels, each of which covers an active area of 100 by 150 $um^2$. The pixels are spread between four barrel layers (BPIX) at 2.9 $<$ r $<$ 16 cm and 3 endcap rings\fxnote{Andrew says it's 'disc', not 'ring'. Each disc is composed of two concentric rings of modules} (FPIX) at 29 $<$ z $<$ 52 cm. The upgraded pixel tracker should provide 4-hit tracking out to eta $<$ 2.4 and improved b-tagging over the original pixel tracker despite the increased pileup expected in 2017.

    Just outside the pixel tracker, the silicon strip tracker (SST) occupies the 20 $<$ r $<$ 116 cm and $|z|$ $<$ 282 cm region of CMS. Composed of 9.3 million silicon strips that vary from \_ to \_ in size \fxnote{look up} depending on particle flux in the region, the SST covers an active area of roughly 196 $m^2$. At each trigger, the strip output voltage is read out via optical link.

\subsection{Electromagnetic Calorimeter}
    \fxnote{cite CMS TDR}
    \fxnote{talk about readout?}
    \fxnote{More in the way of concept/how stuff works?}
    After traversing the inner tracker, a particle would next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses 61200 lead tunstate crystals in the barrel and 7324 in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (80\% of light within 25 ns), compact (radiation length = 0.89 cm), fine-grained (Moliere\fxnote{Andrew says 'radius', not 'length', and use accent!} length = 2.2 cm), and radiation hard (up to 10 Mrad\fxnote{learn context for this number}) calorimeter. The main drawback is the relatively low light yield (30 photon / MeV), which necesitates photodetectors with intrinsic gain that work in magnetic fields \fxnote{say B-field part better}.

    The barrel section (EB) extends radially from 129 to 177 \fxnote{look up} cm and covers a pseudorapidity\fxnote{define} range of 0 to 1.479. The crystals are tapered to approximately project back to the IP but not so perfectly that likely particle trajectories align with cracks. Each crystal is approximately one Moliere radius wide and 25.8 radiation lengths deep.

@@

\subsection{Muon System}
    \fxnote{cite CMS TDR}
    \fxnote{Specifically mention CMS design specs to meet physics goals?}
    The muon system (MS), which is the outermost subdetector, detects muons with 3 varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background, muon rates, and magnetic field strength. The RPCs are useful in the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT\fxnote{define} resolution at high momenta where the tracker pT resolution declines \fxnote{quantify or show std plot?}.

@@

\subsection{Trigger}
    \fxnote{cite CMS TDR, jinst 12 c03021}
    \fxnote{Is entire detector read out at L1?}
    \fxnote{Mention fpgas in L1 or more details about HLT?}
    \fxnote{Sloppy, but I'm getting tired}
    The trigger reduces the data writing rate from the 40 MHz collision rate to less than 1 kHz so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). L1 uses input from ECAL, HCAL, and MS that represents physics objects (e/gamma, tau, jets, MET\fxnote{define}, and mu) and uses custom electronics to reduce the rate to $\approx$100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to further reduce the rate to the desired $<$ 1 kHz.

\section{LHC Searches for Long-lived Particles}
\fxnote{Include diagrams for signal models} 
\fxnote{mention which data was used for each search} 
\fxnote{Steal limit plots and any cool figures from papers} 
\subsection{Overview}
    LLPs leave distinct signatures that differentiate LLP signals from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were generally designed to detect and reconstruct stable particles from prompt decays, so LLP analyses often need to stray from the traditional analysis path in terms of object reconstruction \fxnote{really reconstruction? Maybe object definition?}, background estimation, and which data tier to use \fxnote{data tier is too jargony}. This paper will focus on four complementary LLP signatures: displaced vertices, disappearing tracks, slow-moving particles, and particles that stop completely before decaying.

@@

    Slow and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. Searches for low-velocity particles use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles \fxnote{redundant}, which normally \fxnote{quantify} have beta\fxnote{define} $>$ 0.9. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

    This search uses a simplified model in the split SUSY paradigm as a benchmark. In the simplified model, gluinos are kinematically accessible while squarks are not. The relevant process, then, is pair produced gluinos that decay to a quark and virtual squark, which then decays to a quark and the LSP, which is presumed to be a neutralino \fxnote{steal diagram from andrew's preapproval. I can't access it - ask andrew}. Because the squark mass is so high, the gluino lives long enough to form an R-hadron \fxnote{first use. Define here if not in LLP section} with SM partons and propogate some measureable distance away from the IP before decaying.

@@

    The authors consider two signal processes: (1) electroweak gaugino production resulting in a disappearing track, an ISR jet, and MET and (2) gluino pair production resulting in a disappearing track, four jets, and MET. Process (1) could be the only gaugino production mode at the LHC if the gluino mass is kinematically unaccessable, but process (2), which leads to chargino production in cascade decays, will be dominant if the gluino mass is within reach. The ISR jet in process (1) gives the gauginos something to recoil off of so the MET signal isn't lost \fxnote{My interpretation. Run it by someone who knows stuff. Maybe just leave this out but be prepared to explain it?}.

@@

    This analysis uses the muon system to look for pairs of muons that arrive out of time with bunch crossings. The main backgrounds to such a signal are cosmic muons, beam halo\fxnote{explain}, and muon system noise. Beam halo and muon system noise are easily removed by the selection criteria, but distinguishing signal from cosmic muons requires a more nuanced approach. After generating signal MC and modeling cosmic muons from data obtained in dedicated cosmic runs, the authors found that cosmic muons and signal could be differentiated using time-of-flight information from the muon system RPCs and DTs. Essentially, the direction of muon momentum could be gleaned from the timing information, and the authors required outgoing muons in both the upper and lower hemispheres.\fxnote{Get cartoon from Juliette}

@@

    After estimating the background by extrapolating data from the cosmic-enriched region to the signal region and determining systematic uncertainties related to the MC simulation, trigger acceptance, and luminosity, the authors calculated the expected background for LLP lifetimes ranging from 100 ns to 1 us. In all cases, the expected background was less then one event. Exactly zero events are observed in every lifetime scenario. These results are then interpreted as upper limits on the signal production cross section \fxnote{include limit plots or table}, which are the first limits for stopped particles that decay to muons at the LHC.

@@

    No significant excess over predicted background is observed, so the results are interpreted as upper limits in the production cross section - mass plane for the considered models. By comparing the observed limits and theoretical predictions, the authors also extract mass limits for the gluinos, stops, and staus in various models. \fxnote{Include plots - table is too big and too many numbers to list}

\section{Experimental Viability of Natural SUSY}
    \fxnote{Andrew says: This should really be your most important section. You should try to bring together the results of the searches you talked about (and of other long-lived particle searches and even of more general SUSY searches) and explain what they mean for a possible natural explanation of EWSB. At what point can we claim that there is no natural solution, or at least that there is no natural solution in terms of SUSY? Essentially, nobody cares about what you talked about in Section 6; what's important is that you can give these results a meaningful context in the bigger picture of a natural explanation of EWSB.}

    When building a natural SUSY model, certain constraints are nearly unavoidable\fxnote{Dangerous. What if Linda knows how to avoid them?}. Specifically, the higgsino, which affects mH at tree level; the stop, which affects mH at one loop; and the gluino, which affets mStop at one loop and therefore mH at two, must all be heavy enough to pull mH up to 125 GeV\fxnote{Do I really understand this?} but light enough to avoid excessive fine tuning. Furthermore, the stop and gluino masses are correlated -\fxnote{m-dash?} a large difference between the two would be another source of fine tuning.

@@

    \fxnote{How do LLPs open doors to naturalness and which of those doors have we already closed?} 
    Opening to the possibility of long-lived sparticles increases the parameter space in which natural SUSY could be hiding. LLP analyses such as the ones described in \fxnote{Section whatever}
